---
title: "An introduction to the SingleCellExperiment class"
author: "Davide Risso and Aaron Lun"
date: "Last modified: June 6, 2017; Compiled: `r format(Sys.time(), '%B %d, %Y')`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteEncoding{UTF-8}
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{SingleCellExperiment Vignette}
-->

```{r options, include=FALSE, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(warning=FALSE, error=FALSE, message=FALSE)
```

# Package Status

|                |               |
| -------------- | ------------- |
| Project Status | [![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active) |
| Travis CI      | [![Build Status](https://travis-ci.org/drisso/SingleCellExperiment.svg?branch=master)](https://travis-ci.org/drisso/SingleCellExperiment) |
| Test coverage  | [![Coverage](https://codecov.io/gh/drisso/SingleCellExperiment/branch/master/graph/badge.svg)](https://codecov.io/gh/drisso/SingleCellExperiment) |

# Installation 

This vignette requires the `r Biocpkg("SingleCellExperiment")` and the `r Biocpkg("scRNAseq")` packages.
To install them:

```{r install, eval=FALSE}
BiocInstaller::biocLite("SingleCellExperiment")
BiocInstaller::biocLite("scRNAseq")
```

To load the packages:

```{r load_packages}
library(SingleCellExperiment)
library(scRNAseq)
```

# The `SingleCellExperiment` class

## Definition

The `SingleCellExperiment` class is a light-weight container for single-cell genomics data.
It extends the `RangedSummarizedExperiment` class with the following additional slots:

* `int_elementMetadata` 
* `int_colData`
* `int_metadata`
* `reducedDims`

The `int_` prefix describes _internal_ slots that are not meant for direct manipulation by the user.
Instead, they are set by other user-visible methods, which will be discussed in more detail below.

## Create instances of SingleCellExperiment

There are two main ways to create instances of `SingleCellExperiment`. 
The first is via the constructor:

```{r construct}
counts <- matrix(rpois(100, lambda = 10), ncol=10, nrow=10)
sce <- SingleCellExperiment(assays = list(counts = counts))
sce
```

The second is via coercion from `SummarizedExperiment` objects.

```{r coerce}
se <- SummarizedExperiment(list(counts=counts))
as(se, "SingleCellExperiment")
```

## Setting up a simple example

Here we use a subset of the `allen` data set from the `r Biocpkg("scRNAseq")` package to demonstrate the use of the class.

```{r fluidigm}
data(allen)
allen
```

These data are stored as a `SummarizedExperiment`, so we will first coerce it into a `SingleCellExperiment`.

```{r}
sce <- as(allen, "SingleCellExperiment")
sce
```

# Available methods for the `SingleCellExperiment`

## Adding spike-in information

One of the main additions to `SummarizedExperiment` is the ability for the user to specify the rows corresponding to spike-in transcripts.
This is done with the method `isSpike`, using an appropriate name for the spike-in set.

```{r spikes}
isSpike(sce, "ERCC") <- grepl("^ERCC-", rownames(sce))
sce
```

The identities of the spike-in rows can be easily retrieved using the name of the spike-in set, as shown below.
The names of currently available spike-in sets can also be returned with the `spikeNames` method.

```{r}
table(isSpike(sce, "ERCC"))
spikeNames(sce)
```

While most experimental designs use a single set of spike-ins, the class has the flexibility of including more than one set of spikes.
Let us pretend that the members of the _Adam_ gene family have beeen spiked-in as external genes in these data.

```{r spikes2}
isSpike(sce, "Adam") <- grepl("^Adam[0-9]", rownames(sce))
sce
table(isSpike(sce, "Adam"))
spikeNames(sce)
```

If `isSpike` is used without specifying any name, it will return the union of all spike-in sets.

```{r}
table(isSpike(sce))
```

Note that the `isSpike` and `isSpike<-` methods get and set columns in `int_elementMetadata` and `int_metadata`.
This information is only relevant to package developers and not necessary for routine use of this class.

## Adding size factors

One can also store size factors in the `SingleCellExperiment` object. 
For illustration, we simply compute the total number of reads as size factors here.
Note that more sophisticated methods for computing size factors are available (see, e.g., `r Biocpkg("scran")`).

```{r sizeFactors}
sizeFactors(sce) <- colSums(assay(sce))
head(sizeFactors(sce))
```

We can compute multiple size factors and store them in the object, by providing a name to `sizeFactors`.
This does _not_ affect the values of the unnamed size factors.

```{r sizeFactors2}
sizeFactors(sce, "ERCC") <- colSums(assay(sce)[isSpike(sce, "ERCC"),])
head(sizeFactors(sce, "ERCC"))
head(sizeFactors(sce)) # same as before
```

## Retrieve `colData` and `rowData` information

The `colData` and `rowData` methods can be used to retrieve the stored sample- and gene-level metadata. 
By default, this will only return the user-visible metadata fields, i.e., not including the fields stored in the `int_*` slots.

```{r metadata}
colData(sce)
rowData(sce)
```

However, it is sometimes useful to retrieve a `DataFrame` with the internal fields, i.e., spike-in and size factor information.
This can be achieved by specifying `internal=TRUE`.

```{r metadata2}
colData(sce, internal=TRUE)
rowData(sce, internal=TRUE)
```

See below for some discussion of why an internal storage mechanism is used here.

## Adding low-dimentional representations

For simplicity and speed, we work on a subset of 100 genes. 
To avoid ending up with only uninteresting genes, we extract the 100 genes with maximal variance in the log-transformed counts.

```{r subset}
library(magrittr)
assay(sce) %>% log1p %>% rowVars -> vars
names(vars) <- rownames(sce)
vars <- sort(vars, decreasing = TRUE)

sce_sub <- sce[names(vars[1:100]),]
sce_sub
```

We obtain the PCA and t-SNE representations of the data and add them to the object with the `reducedDims` method.

```{r pca}
library(Rtsne)
set.seed(5252)

pca_data <- prcomp(t(log1p(assay(sce_sub))))
tsne_data <- Rtsne(pca_data$x[,1:50], pca = FALSE)

reducedDims(sce_sub) <- SimpleList(PCA=pca_data$x, TSNE=tsne_data$Y)
sce_sub
```

The stored coordinates can be retrieved by name or by numerical index.
Each row of the coordinate matrix is assumed to correspond to a cell, while each column represents a dimension.

```{r}
reducedDims(sce_sub)
reducedDimNames(sce_sub)
head(reducedDim(sce_sub, "PCA")[,1:2])
head(reducedDim(sce_sub, "TSNE")[,1:2])
```

Any subsetting by column of `sce_sub` will also lead to subsetting of the dimensionality reduction results by cell.

```{r}
dim(reducedDim(sce_sub, "PCA"))
dim(reducedDim(sce_sub[,1:10], "PCA"))
```

## Convenient access to named assays

In the `SingleCellExperiment`, users can assign arbitrary names to entries of `assays`.
To assist interoperability between packages, we provide some suggestions for what the names should be for particular types of data:

- `counts`: Raw count data, e.g., number of reads or transcripts for a particular gene.
- `normcounts`: Normalized values on the same scale as the original counts.
For example, counts divided by cell-specific size factors that are centred at unity.
- `logcounts`: Log-transformed counts or count-like values.
In most cases, this will be defined as log-transformed `normcounts`, e.g., using log base 2 and a pseudo-count of 1.
- `cpm`: Counts-per-million.
This is the read count for each gene in each cell, divided by the library size of each cell in millions.
- `tpm`: Transcripts-per-million.
This is the number of transcripts for each gene in each cell, divided by the total number of transcripts in that cell (in millions).

Each of these suggested names has an appropriate getter/setter method for convenient manipulation of the `SingleCellExperiment`.
For example, we can take the (very specifically named) `tophat_counts` name and assign it to `counts` instead:

```{r}
counts(sce) <- assay(sce, "tophat_counts")
sce
dim(counts(sce))
```

This means that functions expecting count data can simply call `counts()` without worrying about package-specific naming conventions.

# Design decisions

## Package scope

By design, the scope of this package is limited to defining the `SingleCellExperiment` class and some minimal getter and setter methods.
For this reason, we leave it to developers of specialized packages to provide more advanced methods for the `SingleCellExperiment` class.
For example, `r Biocpkg("scater")` defines a number of specific methods such as `normalize` and `r CRANpkg("dplyr")`-like verbs.
We also leave it to the developers of existing packages to provide coercion methods from their classes to `SingleCellExperiment`.

## Why internal storage?

We use an internal storage mechanism to protect the spike-in and size factor fields from direct manipulation by the user.
This ensures that, e.g., only a call to `sizeFactors<-` can change the size factors.
The same effect could be achieved by reserving a subset of columns (or column names) as "private" in `colData()` and `rowData()`, though this is not easily implemented.

The internal storage avoids situations where users or functions can silently overwrite these important metadata fields during manipulations of `rowData` or `colData`.
This can result in bugs that are difficult to track down, particularly in long workflows involving many functions.
It also allows us to add new methods and metadata types to `SingleCellExperiment` without worrying about overwriting user-supplied metadata in existing objects.

## What's up with `reducedDims`? 

We use a `SimpleList` as the `reducedDims` slot to allow for multiple dimensionality reduction results.
One can imagine that different dimensionality reduction techniques will be useful for different aspects of the analysis, e.g., t-SNE for visualization, PCA for pseudo-time inference.
We see `reducedDims` as a similar slot to `assays()` in that multiple matrices can be stored, though the dimensionality reduction results need not have the same number of dimensions.

## Why derive from a `RangedSummarizedExperiment`?

We decided to extend `RangedSummarizedExperiment` rather than `SummarizedExperiment` because for certain assays it will be essential to have `rowRanges()`. 
Even for RNA-seq, it is sometimes useful to have `rowRanges()` and other classes to define the genomic coordinates, e.g., `DESeqDataSet` in the `r Biocpkg("DESeq2")` package.
An alternative would have been to have two classes, `SingleCellExperiment` and `RangedSingleCellExperiment`.
However, this seems like an unnecessary duplication as having a class with default empty `rowRanges` seems good enough when one does not need `rowRanges`.

# Session Info

```{r}
sessionInfo()
```
